---
title: "spida2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spida2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spida2)
```

## spida2: Functions used in the Statistical Consulting Service at York University

The spida2 package is a collection of functions and datasets intended primarily
for statistical consulting, with a particular emphasis on
longitudinal and hierarchical data analysis. Some documents related to this 
package can be found at \url{http://blackwell.math.yorku.ca/R/spida2/doc}

## General Bugs

- `latticeExtra::+.trellis()` in xyplot + xyplot requires the same data set to avoid registration of panels if some levels of panel factors are different in data sets for different panes. 
- Use `Rbind()` to create a common data frame and different names for plotted tables, e.g. to plot both lines and points. 
}

## New functions

-  `Rbind()` works on data.frame, missing variables get NA: use to avoid fted panels in latticeExtra
-  `Apply()` returns list with same structure, e.g. list array
-  `tabf()` returns result of applying function as a list array
-  `waldx()` temporary name for rank-deficient aware version of wald
-  `waldf()` rank-deficient aware version of wald that returns a data frame  L matrix
-  `subrow()` subtracts selected rows of L matrix from ranges of rows, e.g. compare with comparators
-  `lchol()` returns lower-triangular L so that G = L'L.
-  `getR()`, `getG()` and `getV()` urn R, G and V matrices for `nlme::lme()` objects.
-  `pdInd()` constructs a pdClass for a G matrix with patterns of zero ariances. See
a vignette at [pdInd: G matrix with a pattern of zeros](http://130.63.179.181/R/spida2/doc/pdInd.html).
}

## Wald tests and linear hypothesis matrices

-  `wald()` Wald tests with L matrices optionally created with regular ressions. 
    Uses SVD to handle linear dependencies in rows of L
-  `walddf()` version of wald that returns a data frame
-  `as.data.frame.wald()` return a data frame from a wald object
-  `coef.wald()` method to extract estimated coefficients
-  `print.wald()` printing method
-  `Lfx()`  creates hypothesis matrices for derivatives and differences.  example for factor differences
-  `M()` constructor for M objects to generate portions of design and othesis matrices. Used with  `Lfx()`
-  `rpfmt()` format estimated values and p-values from a wald test
-  `Lall()` for lmer objects
-  `Lc()` for lmer objects
-  `Lmu()` for lmer objects
}

## Utilities for fitted objects

-  `getD()` get data frame from a fitted object
-  `getData()` older version using methods. Might work if previous fails
-  `getFix()` get fixed effects from a fitted object
-  `getX()` get X matrix from fitted object
-  `getV()` get V matrix from a mixed model
-  `getG()` get G matrix from a mixed model
-  `getR()` get R matrix from a mixed model
-  `Vcov()` get estimated variance covariance of fixed effects from a ted object
}

## Multilevel data frames

-  `capply()`: `capply(x,id,FUN)` applies the
  function `FUN` to chunks of 'x' formed by levels of 'id'.
  The result has the same form as 'x' with replication within
  chunks, if needed.
-  `up()`, `agg()` and `up_apply()` create summary
  data sets consisting, by default, of within-id-invariant variables. 
  Summaries of id-varying variables can also be included. `agg()` can
  create mean incidence matrices for lower-level factors.
-  `cvar()` and `dvar()` are designed to be used
  in linear model formulas to generate 'centered-within-group' and 'within-group
  deviation' variables. WIth factors, they generate mean incidence matrices.
-  `varLevel()` and `gicc()`: the level of a variable h respect
  to a clustering formula and the 'generalized' intra-class correlation coefficient.
-  `tolong()` and `towide()` are 
  interfaces to stats(reshape) to facilitate
  the typical uses of reshape for longitudinal data.
}

## Splines -- parametric and non-parametric

-  `gsp()` creates a function for a generalized 
      spline that can
      be included in a linear model formula. 
-  `sc()` creates a spline contrast matrix for 
      general spline hypotheses. The matrix can be included
      in hypothesis matrices for the `wald()` function.
-  `smsp()` creates a matrix for a smoothing spline.
}

## Datasets

-  `hsfull()` Classical data set on high school math achievement and ses.  Bryk and Raudenbush and many other sources.
-  `iq()` Recovery after traumatic brain injury
-  `Drugs()` Longitudinal data on drugs and schizophrenia symptoms. ustrates role of control variables with non-random assignment. 
-  `Indonesia()` Xerophthalmia.
-  `migraines()` Longitudinal data on migraine treatment and weather
-  `coffee()` Artificial data on coffee, heart damage and stress. ustrates Simpson's Paradox with continuous predictors.
-  `hw()` Artificial data on height, weight and health. Illustrates pression.
-  `Unemp()` U.S. monthly unemployment from January 1995 to February 2019.
}

## Graphics

-  `gd()` and `td()` are easy interface to set
      graphical parameters for lattice and graphics. `gd()` sets
      parameters to make graphs look like ggplot2 graphics. 
-  `panel.fit()` add fitted values and error bands with `latticeExtra::layer()` or `latticeExtra::glayer()`
-  `panel.dell()` adds data ellipse with `latticeExtra::layer()` or `latticeExtra::glayer()`
}

## Miscellaneous utility functions

-  `sortdf()` sort rows of a data frame -- useful in a magrittr pipeline
-  `assn()` assign -- useful in a magrittr pipeline
-  `disp()` utility to display value of a variable -- useful for debugging
-  `getFactorNames()` get names of variables that are factors in a data me
-  `%less%()` synonym for `setdiff()` as well as
      `%and%()` and `%or%()`
-  `labs()` assign, extract and print labels for various objects
-  `pch()` generate plotting character mnemonically
-  `pfmt()` format p-values
-   `print.cat()`
-  `rnd()` round a vector to keep significant digits in variation in values
-  `run()` evaluate a string as a command with try
-  `grepv()` grep(..., value = TRUE)
}

## String manipulation functions

Function designed to work smoothly with pipes in \pkg{magrittr}:
-  `sub_()` and `gsub_()` handle substitution inline and return a factor if the input is a factor. 
-  `name()` changes the names of an object and returns the renamed object 

